---
title: "Introduction to the quakeR package"
author: "Vadim Bondarenko"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the quakeR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Intro

The package `quakeR` is designed for working with the NOAA Significant Earthquakes dataset. The dataset has a substantial amount of information that is not immediately accessible to people without knowledge of the intimate details of the dataset or of R. This package provides the tools for processing and visualizing the data so that others may extract some use out of the information embedded within.

## Installaton

```{r}
library(quakeR)
```


## The Data

This project is centered around a dataset obtained from the U.S. National Oceanographic and Atmospheric Administration (NOAA) on significant earthquakes around the world. This dataset contains information about 5,933 earthquakes over an approximately 4,000 year time span.

The data set can be downloaded from this [link](https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1). Alternatively, a copy of the dataset is included with this package, and can be loaded in R with the following code:

```{r message=FALSE}
filename <- system.file("extdata/earthquakes.tsv.gz", package = "quakeR")
raw_data <- readr::read_delim(filename, delim = "\t")
```




## Functions

### Clean Dats

Takes raw NOAA data frame and returns a clean data frame. The clean data frame has the following: A date column created by uniting the year, month, day and converting it to the Date class LATITUDE and LONGITUDE columns converted to numeric class.

```{r}
library(dplyr, warn.conflicts = FALSE)

clean_data <- eq_clean_data(raw_data)
clean_data %>% 
  arrange(desc(DATE)) %>% 
  select(DATE, COUNTRY, LOCATION_NAME, LATITUDE, LONGITUDE, DEATHS) %>% 
  head()
```

